---
title: "Reference gene stability evaluation: AquaFly1-PI"
author: "Yanxian Li"
date: "`r Sys.time()`"
output:
  html_document: 
    self_contained: yes
    toc: true
    number_sections: true
  pdf_document: 
    latex_engine: xelatex
    toc: true
    number_sections: true
  word_document: default
---

```{r setup, include=FALSE}
library(here) # a replacement for 'file.path()'

library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, comment=NA, prompt=FALSE, cache=FALSE, dev='tiff', dev.args=list(compression='lzw'), dpi=300, fig.path=here("figs", "qPCR_ref", "ref_eval_"))

# The following configurations are necessary for dfSummary() to work properly
library(summarytools)
st_options(plain.ascii = FALSE,          # This is very handy in all Rmd documents
           style = "rmarkdown",          # The same
           footnote = NA,                # Avoids footnotes which would clutter the results
           subtitle.emphasis = FALSE,    # According to the theme used, might improve the headings' layout
           dfSummary.varnumbers = FALSE, # This keeps results narrow enough
           dfSummary.valid.col = FALSE)  # The same
```

```{r, include=FALSE}
st_css() # This is a must; without it, expect odd layout, especially with dfSummary()                    
```

# Data import and tidy 
## Import data
```{r}
library(tidyverse) # tidy, manipulate and plot data
df <- read_csv(here("data" ,"AqFl2_qPCR_ref_PI.csv"), col_names = T, na = "") 
str(df)
```

## Tidy data
```{r}
# Tidy data
df <- df %>% 
  arrange(Ref_gene, desc(Diet)) %>% 
  mutate(Quantity = `^`(PE, -Cq)) %>% # calculate mRNA quantity
  group_by(Ref_gene, Diet)  %>%
  mutate(Mean = mean(Quantity), SD = sd(Quantity)) %>% # calculate average mRNA quantity and sd
  ungroup() %>%
  group_by(Ref_gene) %>%
  mutate(Quantity_std = scale(Quantity), # standardize mRNA quantity of each reference gene
         Quantity_nm = Quantity / Quantity[1]) # normalize mRNA quantity to the first sample

# Convert Sample_ID and Net_pen to character variables
df <- within(df, {
  Sample_ID <- as.character(Sample_ID)
  Net_pen <- as.character(Net_pen)
})

# Convert Diet and Sample_ID to factors, specifying the desired orders for plotting
df$Diet <- factor(df$Diet, levels = c("REF", "IM"))
df$Sample_ID <- factor(df$Sample_ID, 
                       levels = unique(df$Sample_ID)) # use the exact order of Sample_ID column 
```

# Exploratory data analysis (EDA) 
## EDA by descriptive statistics 
As packages providing descriptive statistics summarize data by columns, we'll convert the dataframe from the "long" to "wide" using the dcast() function from the *data.table* package. 
```{r}
library(data.table)
# Convert dataframe to a data.table object
dt <- as.data.table(df)

# Transform data.table from long to wide
df_wider <- dcast(dt, Sample_ID + Diet + Net_pen ~ Ref_gene, value.var = colnames(dt)[5:13])
```

### Descriptive statistics with *summarytools*
```{r summarytools, results = 'asis'} 
library(summarytools) 
print(dfSummary(df_wider, valid.col = FALSE, graph.magnif = 0.75), max.tbl.height = 800, method = "render")
```

### Descriptive statistics with *skimr*
```{r skimr, results = 'asis'}
library(skimr)
library(kableExtra) # Format table

skim_to_wide(df_wider) %>%
  kable(align = "l") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## EDA by visualization
```{r}
library(cowplot) #  a ggplot2 add-on to provide a publication-ready theme 
```

### Dot chart
```{r dot_chart, fig.height=10, fig.width=8}
# Make a dataframe containing the grand mean of mRNA quantity 
gm <- df %>% 
  group_by(Ref_gene) %>%
  summarize(gmean = mean(Quantity))

# Make dot charts showing individual expression profile
p1 <- df %>% 
  ggplot(aes(x = Sample_ID, y = Quantity, group = 1)) +
    geom_point(aes(color = Diet)) +
    geom_line() +
    facet_wrap(~ Ref_gene, ncol = 1, scales = "free_y") +
    labs(y = "mRNA quantity", tag = "A") +
    scale_y_continuous(labels = scales::scientific) + 
    geom_hline(aes(yintercept = gmean), gm, color = "blue") + # The blue line shows the grand mean.
    theme_bw() 

# Make dot charts showing normalized expression levels of all the reference genes in one plot
p2 <- df %>% 
  ggplot(aes(x = Sample_ID, y = Quantity_nm, color = Ref_gene)) +
    geom_point() +
    geom_line(aes(group = Ref_gene)) +
    labs(y = "normalized mRNA quantity", tag = "B") +
    theme_bw() +
    scale_fill_brewer(palette = "Dark2") 

# Combine the plots
plot_grid(p1, p2, ncol = 1, align = "v", rel_heights = c(3, 2))
```

### Boxplot overlaying violin plot
```{r boxplot}
# Convert "Sample_ID"" to character, otherwise ggrepel won't work properly
df$Sample_ID <- as.character(df$Sample_ID)

# Define a function for identifying outliers
is_outlier <- function(x) {
  x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x)
}

# Make plots using standardized mRNA quantity
set.seed(1910)

df %>% 
  group_by(Ref_gene) %>%
  mutate(outlier = is_outlier(Quantity_std)) %>%
  ggplot(aes(x = Ref_gene, y = Quantity_std, label = ifelse(outlier, Sample_ID, NA))) +
  geom_violin(trim = T) +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_point(aes(fill = Diet, colour = Diet), size = 3, shape = 21, position = position_jitterdodge(0.2)) +
  ggrepel::geom_text_repel(aes(colour = Diet), position = position_jitterdodge(0.2)) + # label outliers
  ylab("standardized mRNA quantity") +
  theme_cowplot() +
  guides(colour = F)
```

### Bar plot
```{r barplot}
df %>%
  ggplot(aes(x = Diet, y =  Mean, fill = Diet)) +
  geom_bar(stat = "identity", position = position_dodge(), colour = "black") +
  geom_errorbar(aes(ymin = Mean, ymax = Mean + SD), size = 0.3, width = 0.2) + # add error bar (sd)
  facet_wrap(~ Ref_gene, nrow = 1, scales = "free_y") +
  scale_y_continuous(limits = c(0, NA), 
                     expand = expand_scale(mult = c(0, 0.1))) + 
  ylab("mRNA quantity") +
  theme_cowplot()
```

### Correlation 
Make a subset of data for correlation analysis
```{r}
# Subset dataframe and convert to the wide format
df1 <- df %>%
  select(Sample_ID, Ref_gene, Quantity) %>%
  spread(key = "Ref_gene", value = "Quantity")
```

#### Visualize correlation matrix using *PerformanceAnalytics*::chart.Correlation()
An excellent implementation of correlation matrix visualization is from the *PerformanceAnalytics* package. 

```{r cor_chart}
library("PerformanceAnalytics")
chart.Correlation(df1[2:ncol(df1)], histogram = TRUE, pch = 19)
```

#### Visualize correlation matrix using *corrplot*::corrplot()
```{r cor_plot}
# Compute a correlation matrix
corr <- cor(df1[2:ncol(df1)], method = "pearson", use = "complete.obs")

# Make a correlation plot
library(corrplot)
corrplot.mixed(corr, order = "hclust", tl.col = "black")
```

### heatmap
```{r}
# Make a numeric metrix for plotting heatmap
mat <- as.matrix(df1[2:ncol(df1)]) %>% 
  `row.names<-`(df1[[1]]) # Sample_ID as rownames

# Define color scheme
col <- colorRampPalette(c("blue", "white", "red"))(25)
```

#### Static heatmap
```{r static_heatmap, fig.height=5}
library(gplots)
heatmap.2(mat,                     
          Rowv = TRUE,
          Colv = TRUE,
          distfun = function(x) dist(x,method = 'euclidean'),
          hclustfun = function(x) hclust(x,method = 'ward.D2'),
          ylab = "Sample ID",
          dendrogram = "column",
          density.info = "none",
          scale = "column",
          trace = "none",
          cexCol = 1,
          cexRow = 0.5,
          srtCol = 45,
          col = col)                       
```

#### Interactive heatmap
```{r interactive_heatmap, fig.height=1.5, fig.width=2}
library(d3heatmap)
d3heatmap(mat,                     
          Rowv = TRUE,
          Colv = TRUE,
          distfun = function(x) dist(x,method = 'euclidean'),
          hclustfun = function(x) hclust(x,method = 'ward.D2'),
          dendrogram = "column",
          density.info = "none",
          scale = "column",
          trace = "none",
          col = col)
```

# Evaluation of reference gene stability
## Ranks by coefficient of variance (cv)
Coefficient of variance (cv) is a measurement of overall stability of reference genes across all the samples. The smaller the CV is, the better.
```{r rank_by_cv}
cv <- df %>% 
  group_by(Ref_gene) %>%
  summarize(Cq_max = max(Cq), 
            Cq_min = min(Cq), 
            Cq_range = max(Cq) - min(Cq), 
            Quantity_mean = mean(Quantity), 
            Quantity_SD = sd(Quantity),
            Quantity_CV = 100 * Quantity_SD / Quantity_mean) %>%
  mutate(CV_rank = dense_rank(Quantity_CV)) %>% # rank reference genes by CV
  arrange(Ref_gene)
```

## Ranks by F-statistic
The F-statistic is a measurement of reference gene stability across experimental treatments. The smaller the F-statistic is, the better.
```{r rank_by_F}
library(broom) # tidy statistical outputs
F_stat <- df %>% 
  group_by(Ref_gene) %>% 
  do(tidy(aov(Quantity ~ Diet, data = .))) %>% # run ANOVA and tidy statistical outputs
  na.omit() %>%
  select(Ref_gene, statistic) %>%
  rename(Quantity_F = statistic) %>%
  as.data.frame() %>%
  mutate(F_rank = dense_rank(Quantity_F)) %>% # rank reference genes by F statistic
  arrange(Ref_gene) 
```

## Summary of ranks
```{r summary}
# Merge tables
smr <- full_join(cv, F_stat, by = "Ref_gene")

# Format digits
smr$Quantity_mean <- formatC(smr$Quantity_mean, format = "e", digits = 1)
smr$Quantity_SD <- formatC(smr$Quantity_SD, format = "e", digits = 2)

# Format table
kable(smr, format = "html", digits = 2, align = "l") %>%
      kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14, position = "left")
```

# Session info
```{r, echo=FALSE}
# Get session info
sessionInfo()
```
